### 2024-04-22 chap_methodo/chap_methodo.tex
RobPol + Methodo PCA: Review by HJ

\subsection{Principal components of the transition}
\label{subsec:meth:PCA:transition}
As introduced, the objective is to define the main technological drivers of the variation through the transition to 2050 subject to uncertainties. To do so, before calculating the PCs of the transition, three preliminary steps are necessary: (i) selection of the right data, (ii) data scaling and, (iii) outliers management. Like any other dimension-reduction process, \gls{PCA} has to be supplied with relevant data to reach the stated objective.  Like the normalization of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude. Finally, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases. After this preprocessing, PCs can be computed for each of the year of the transition then aggregated to give a bigger picture over the whole transition.\\

@@

\noindent
To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})). Even though these represent only the design part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:c_inv})) and, more indirectly, the resources to use.  In other words, it captures the technological landscape of the transition.  Having defined the type of variable to consider, we need to assemble a relevant dataset. This is given by the runs to quantify the impact of the uncertain parameters on the total cost of transition required by the method described in Section \ref{sec:meth:UQ}. Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ stands for the sample number of the uncertainty quantification method. For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties. Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}). In other words, we focus here only on the years 2025, 2030, 2035, 2040, 2045 and 2050. This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}). Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.\\

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Share_EUD_sectors.pdf}
\caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition. These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimized through the perfect foresight approach. Over the transition, sectors like electricity (\ie from 22\% in 2025 to 33\% in 2050) or ammonia (\ie from 2\% to 8\%) become more important due to sector coupling, \eg e-mobility or ammonia-\gls{CCGT}.}
\label{fig:Share_EUD_sectors}
\end{figure}

To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal value (see Chapter \ref{chap:atom_mol}).  The end-use-demands as well as the commodity produced for the sector coupling are based on the results of this deterministic REF case.  For instance, the share of the electricity sector accounts for its \gls{EUD} and the electricity produced to supply other sectors (\eg heat, mobility). Finally, to compare apples with apples, we converted the \gls{EUD} in the mobility sectors, \ie passenger and freight, into the \gls{FEC} they require in the REF case. This gives the second weighing factor to scale data, on top of the one of Equation (\ref{eq:PCA_scaling_0}) (see Equation (\ref{eq:PCA_scaling_1})).

@@

One would notice that this second scaling factor omits the infrastructure and storage technologies. In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact. Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies (\eg \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}). \\

@@

\noindent
Handling outliers is one of the biggest challenges in data science \cite{aguinis2013best}. These are defined as data points differing significantly from the rest of the data set. Being extreme values, outliers influence the overall dataset variance and, consequently, rotate the PCs directions towards them \cite{stanimirova2007dealing}. In the context of \gls{PCA}, outliers could be defined as ``model fit outliers'' as their presence influences the fit of the model. There are several techniques to detect/define and handle the outliers \cite{aguinis2013best}. In this work, detection is performed via the box plot technique, as outliers are identified as those points lying beyond the plot’s whiskers, or fences. These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}). Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the rest of the dataset for this specific technology and year across all the samples. There exist several techniques to handle these outliers depending on their nature or the method used to identify them. Since all the data points correspond to a result provided by the optimization, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}. In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively. In practice, this modification narrows the variation range of features presenting outliers and, consequently, reduces their weight in the different PCs.\\

@@

\noindent
Now that data are selected and preprocessed, principal components are first computed for each representative year separately, using the Python package \textsc{PCA} from \textsc{sklearn.decomposition}. As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.} in our case study) which is intractable. Moreover, the first PCs keep track of most of the variance of the system whereas the last ones present a smaller interest. Choosing the appropriate threshold involves a trade-off. Retaining too few principal components may result in loss of important information, while retaining too many may lead to unclear analysis. In this work is to compute, for each of the representative years of the transition (except 2020),  the PCs capture 90\% of the total variance of this year \cite{jolliffe2002principal}. At the end of this step, we have a list of $m$ PCs, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and

@@

\noindent
where $m$ is presumably different for each representative year and $p$ is the total number of variables, hence the maximum number of PCs. For the entire transition, it gives a total of $M$ $\text{PC}_{y,i}$.\\

@@

\noindent
The final step consists in defining metrics on the whole transition based on the $\text{PC}_{y,i}$ computed for each representative year, separately. To do so, all the $\text{PC}_{y,i}$ from every year are sorted together in a descending order based on their respective variance. Then, starting with the one with the highest absolute variance, all the other $\text{PC}_{y,i}$ similar to it are clustered together. The similarity between two PCs is defined according to the cosine similarity approach, especially appropriate in high-dimensional positive spaces \cite{xia2015learning}. Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest. Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Equation (\ref{eq:cos_similarity})). It is important to consider to second specific case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors. For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$. The second vector being the opposite of the first, they are considered as equivalent in the regard of \gls{PCA}.

@@

\noindent
where $A$ and $B$ represent two different PCs.  The components of these similar PCs are then averaged to form the first PC of the transition, $\text{PC}_{\text{transition},1}$. Then, the process repeats with the $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$. This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to $\alpha$\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$, \ie the ``total transition variance'', generated at the previous step:

@@

In our case, we have decided to set the value of $\alpha$ to 80\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{y}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}). For instance, capturing up to 90\% of the the total transition variance would mean almost doubling the number of $\text{PC}_{y}$ to consider and, consequently, the number of directions of variation to assess.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.5\textwidth]{PC_y_share_tot_var.pdf}
\caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{y}$. Passing a certain threshold, here 80\%, the number of $\text{PC}_{y}$ to consider increase significantly compared to the share of transition variance that they capture.}
\label{fig:PC_y_share_tot_var}
\end{figure}

@@

\noindent
Similarly to the work of \citet{moret2020overcapacity}, roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimizations. The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps (see Figure \ref{fig:Methodo_Robustness}).

>>>>

\subsection{Principal components of the transition}
\label{subsec:meth:PCA:transition}
As introduced, the objective is to define the main technological drivers of the variation through the transition to 2050 subject to uncertainties. To do so, before calculating the PCs of the transition, three preliminary steps are necessary: (i) selection of the right data, (ii) data scaling and, (iii) outliers management. Like any other dimension-reduction process, \gls{PCA} has to be supplied with relevant data to reach the stated objective.  Like the normalization of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude. Then, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases. After this preprocessing, PCs can be computed for each of the representative year of the transition. Finally, PCs that indicate similar directions in different years are averaged into one single PCs of transition. The different PCs of transition give a bigger picture over the whole transition.\\

@@

\noindent
To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})). Even though these represent only the design part of the result of the optimisation focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation. In other words, it captures the technological landscape of the transition.  Having defined the type of variable to consider, we need to assemble a relevant dataset. This is given by the runs to quantify the impact of the uncertain parameters on the total cost of transition required by the method described in Section \ref{sec:meth:UQ}. Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ stands for the sample number of the uncertainty quantification method. For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties. Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}. Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}). In other words, we focus here only on the years 2025, 2030, 2035, 2040, 2045 and 2050. This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}). \\

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=1\textwidth]{Share_EUD_sectors.pdf}
\caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition. These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimized through the perfect foresight approach. Over the transition, sectors like electricity (\ie from 22\% in 2025 to 33\% in 2050) or ammonia (\ie from 2\% to 8\%) become more important due to sector coupling, \eg e-mobility or ammonia-\gls{CCGT}.}
\label{fig:Share_EUD_sectors}
\end{figure}

To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal values (see Chapter \ref{chap:atom_mol}).  The end-use-demands as well as the commodity produced for the sector coupling are based on the results of this deterministic REF case.  For instance, the share of the electricity sector accounts for its \gls{EUD} and the electricity produced to supply other sectors (\eg heat, mobility). Finally, to compare apples with apples, we converted the \gls{EUD} in the mobility sectors, \ie passenger and freight, into the \gls{FEC} they require in the REF case. This gives the second weighing factors to scale data, on top of the ones of Equation (\ref{eq:PCA_scaling_0}), see Equation (\ref{eq:PCA_scaling_1}).

@@

One can notice that this second scaling factor omits the infrastructure and storage technologies. In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact. Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies (\eg \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}). \\

@@

\noindent
Handling outliers is one of the biggest challenges in data science \cite{aguinis2013best}. These are defined as data points differing significantly from the rest of the data set. Being extreme values, outliers influence the overall dataset variance and, consequently, rotate the PCs directions towards them \cite{stanimirova2007dealing}. In the context of \gls{PCA}, outliers could be defined as ``model fit outliers'' as their presence influences the fit of the model. There are several techniques to detect/define and handle the outliers \cite{aguinis2013best}. In this work, detection is performed via the box plot technique, as outliers are identified as those points lying beyond the plot’s whiskers, or fences. These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}). Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for this specific technology and year. There exist several techniques to handle these outliers depending on their nature or the method used to identify them. Since all the data points correspond to a result provided by the optimization, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}. In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively. In practice, this modification narrows the variation range of features presenting outliers and, consequently, reduces their weight in the different PCs.\\

@@

\noindent
Now that data are selected and preprocessed, principal components are first computed for each representative year separately, using the Python package \textsc{PCA} from \textsc{sklearn.decomposition}. As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73 in our case study\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.}) which is intractable. Moreover, the first PCs keep track of most of the variance of the system whereas the last ones present a smaller interest. Choosing the appropriate threshold involves a trade-off. Retaining too few principal components may result in loss of important information, while retaining too many may lead to unclear analysis. For each of the representative years of the transition (except 2020), we compute the PCs that capture 90\% of the total variance of this year \cite{jolliffe2002principal}. At the end of this step, a list of $m$ PCs is obtained, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and

@@

\noindent
where $m$ is presumably different for each representative year and $p$ is the total number of variables, hence the maximum number of PCs. For the entire transition, it gives a total of $M=34$ $\text{PC}_{y,i}$.\\

@@

\noindent
The final step consists in defining metrics on the whole transition based on the $\text{PC}_{y,i}$ computed for each representative year, separately. To do so, all the $\text{PC}_{y,i}$ from every year are sorted together in a descending order based on their respective variance. Then, starting with the one with the highest absolute variance, all the other $\text{PC}_{y,i}$ similar to it are clustered together. The similarity between two PCs is defined according to the cosine similarity approach, especially appropriate in high-dimensional positive spaces \cite{xia2015learning}. Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest. Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Equation (\ref{eq:cos_similarity})). It is important to consider the second case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors. For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$. The second vector being the opposite of the first, they are considered as equivalent in the regard of \gls{PCA}.

@@

\noindent
where $A$ and $B$ represent two different PCs.  The components of these similar PCs are then averaged to form the first PC of the transition, $\text{PC}_{\text{transition},1}$. Then, the process repeats with the next $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$. This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to $\alpha$\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$, \ie the ``total transition variance'', generated at the previous step:

@@

In our case, we have decided to set the value of $\alpha$ to 85\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{\text{transition}}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).  Besides being the metrics to assess the robustness of roadmap, these $\text{PC}_{\text{transition}}$ also point out the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.5\textwidth]{PC_y_share_tot_var.pdf}
\caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{\text{transition}}$. Passing a certain threshold, here 85\%, the number of $\text{PC}_{\text{transition}}$ to consider increase significantly compared to the share of transition variance that they capture.}
\label{fig:PC_y_share_tot_var}
\end{figure}

@@

\noindent
In their analysis, \citet{moret2020overcapacity} assessed the additional capacities after the realisation of uncertainties, the so-called \textit{possibility of recourse} \cite{grossmann2016recent}. These adaptations from the initial investments lead to overcapacity where assets are built but not used because the values taken by uncertain parameters make them cost inefficient. To address the same question of overcapacity in the context of energy transition pathway with progressively revealed uncertainties, we have adapted the method of \citet{moret2020overcapacity} (see Figure \ref{fig:Methodo_Robustness}). Roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimisations. The $\text{PC}_{\text{transition}}$ are the metrics on which the results from the myopic pathway optimisation subject to minimal installed capacities set by these roadmaps are projected.
